---
title: "Variability in socioeconomic status among different zip codes in NYC."
authors: "Limor Kohanim & Sushobhan Parajuli"
date: "6/22/2022"
output: html_document
---
In this document, we will extract data for six different socioeconomic status from United  States  Census  Bureau and plot them for all the zip-codes in New York City.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will need an api key to access the census data. Here is our api key.

API: a1dc3b91bf844ea51a5e86368c659cb801e081f6
```{r}
#census_api_key("a1dc3b91bf844ea51a5e86368c659cb801e081f6", install = TRUE)
#readRenviron("~/.Renviron")
```

First, let's import the libraries we will need.
```{r}
library(tidycensus)
library(tidyverse)
library(tigris)

options(tigris_use_cache = TRUE)
library(ggplot2)
library(tidyr)
```

Now, we will import the list of all the zip codes in NYC. "nyc_zip_codes" is a csv file that has a column named "ZipCode" that contains all NYC zip codes. This data is available on github, on the same link we use here.
```{r}
nyc_zip_codes <- read_csv(
  "https://raw.githubusercontent.com/erikgregorywebb/nyc-housing/master/Data/nyc-zip-codes.csv",
  show_col_types = FALSE)
```

Let's examine the plots. Before each codes will examine sub codes that we need with load_variables.
---

Following code plots a heat map for proportion of the 18-64 year olds who are uninsured across all the zip codes in NYC.
```{r}
#v1 <- load_variables(2016, "acs5") %>% filter(grepl("B27010", name))

uninsured <- get_acs(geography = "zcta",
                     variables = c('B27010_018','B27010_034','B27010_033', 'B27010_050'),
                     state = 'NY', 
                     year = 2016, 
                     survey = 'acs5', 
                     geometry = TRUE) %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) %>%
  mutate(prop = (estimate_B27010_033+estimate_B27010_050)/
           (estimate_B27010_034+estimate_B27010_018))

uninsured <- uninsured[uninsured$GEOID %in% nyc_zip_codes$ZipCode, ]

plot_1 <- ggplot(data = uninsured, 
       mapping = aes(fill = uninsured$prop)) + 
  geom_sf(data = uninsured$geometry, color = "gray", lwd = 0.1) + 
  scale_fill_distiller(palette = "YlOrRd", direction = 1) + 
  theme_void() +
  labs(title = "Proportion of 18-64 who are uninsured",
       fill = "") +
  theme(plot.title = element_text(size=8))
```

Following code plots a heat map for median income across all the zip codes in NYC.
```{r}
#v2 <- load_variables(2016, "acs5") %>% filter(grepl("B19013", name))

median_income <- get_acs(geography = "zcta",
                         variables = 'B19013_001',
                         state = 'NY',
                         year = 2016, 
                         survey = 'acs5',
                         geometry = TRUE)

median_income <- median_income[median_income$GEOID %in% nyc_zip_codes$ZipCode, ]

plot_2 <- ggplot(data = median_income, 
       mapping = aes(fill = median_income$estimate / 1000000)) + 
  geom_sf(data = median_income$geometry, color = "gray", lwd = 0.1) + 
  scale_fill_distiller(palette = "YlGn", direction = 1) + 
  theme_void() +
  labs(title = "Median income ( in millions, 2016$)",
       fill = "") +
  theme(plot.title = element_text(size=8))

```

The next code plots a heat map for proportion of self-identifying whites across all zip codes in NYC.
```{r}
# v3 <- load_variables(2016, "acs5") %>% filter(grepl("B02001", name))

selfid_whites <- get_acs(geography = "zcta",
                 variables = c('B02001_001', 'B02001_002'),
                 state = 'NY', 
                 year = 2016, 
                 survey = 'acs5', 
                 geometry = TRUE) %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) %>%
  mutate(prop = estimate_B02001_002/estimate_B02001_001)

selfid_whites <- selfid_whites[selfid_whites$GEOID %in% nyc_zip_codes$ZipCode,]

plot_3 <- ggplot(data = selfid_whites,
       mapping = aes(fill = selfid_whites$prop)) + 
  geom_sf(data = selfid_whites$geometry, color = "gray", lwd = 0.1) + 
  scale_fill_distiller(palette = "Purples", direction = 1) + 
  theme_void() +
  labs(title = "Proportion self-identifying as White",
       fill = "") +
  theme(plot.title = element_text(size=8))

```

The next code plots a heat map for proportion in households of 4 or more across all zip codes in NYC. 
```{r}
#v4 <- load_variables(2016, "acs5") %>% filter(grepl("B11016", name))

households4 <- get_acs(geography = "zcta", 
                      variables = c('B11016_001', 'B11016_005', 'B11016_006', 'B11016_007',
                                    'B11016_008',
                                    'B11016_013', 'B11016_014', 'B11016_015', 'B11016_016'),
                      state = 'NY', year = 2016, geometry = TRUE) %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) %>%
  mutate(prop = (estimate_B11016_005+estimate_B11016_006+
                                      estimate_B11016_007+estimate_B11016_008+
                                      estimate_B11016_013+estimate_B11016_014+
                                      estimate_B11016_015+estimate_B11016_016)/estimate_B11016_001)

households4 <- households4[households4$GEOID %in% nyc_zip_codes$ZipCode,]

plot_4 <- ggplot(data = households4,
       mapping = aes(fill = households4$prop)) +
  geom_sf(data = households4$geometry, color = "gray", lwd = 0.1) + 
  scale_fill_distiller(palette = "YlOrRd", direction = 1) + 
  theme_void() +
  labs(title = "Proportion in households of 4 or more",
       fill = "") +
  theme(plot.title = element_text(size=8))

```

Following code plots a heat map for proportion of population that commutes by bus across all zip code in NYC.
```{r}
#v5 <- load_variables(2016, "acs5") %>% filter(grepl("B08301", name))

commutebybus <- get_acs(geography = "zcta", 
                      variables = c('B08301_001', 'B08301_011'),
                      state = 'NY', year = 2016, geometry = TRUE) %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) %>%
  mutate(prop = estimate_B08301_011 / estimate_B08301_001)

commutebybus <- commutebybus[commutebybus$GEOID %in% nyc_zip_codes$ZipCode,]

plot_5 <- ggplot(data = commutebybus,
       mapping = aes(fill = commutebybus$prop)) +
  geom_sf(data = commutebybus$geometry, color = "gray", lwd = 0.1) + 
  scale_fill_distiller(palette = "YlOrRd", direction = 1) + 
  theme_void() +
  labs(title = "Proportion of population that commutes by bus",
       fill = "") +
  theme(plot.title = element_text(size=8))
```

And this final code plots a heat map for proportion of population 65+ years of age across all zip codes in NYC.
```{r}
#v6 <- load_variables(2016, "acs5") %>% filter(grepl("B01001", name))

pop65andabove <- get_acs(geography = "zcta", 
                      variables = c('B01001_001', 'B01001_020', 'B01001_021', 'B01001_022',
                                        'B01001_023','B01001_024', 'B01001_025', 'B01001_044',
                                        'B01001_045', 'B01001_046', 'B01001_047',
                                        'B01001_048', 'B01001_049'),
                      state = 'NY', year = 2016, geometry = TRUE) %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) %>%
  mutate(prop = (estimate_B01001_020+estimate_B01001_021+estimate_B01001_022+estimate_B01001_023+
           estimate_B01001_024+estimate_B01001_025+estimate_B01001_044+ estimate_B01001_045+
           estimate_B01001_046+estimate_B01001_047+estimate_B01001_048+estimate_B01001_049)/estimate_B01001_001)

pop65andabove <- pop65andabove[pop65andabove$GEOID %in% nyc_zip_codes$ZipCode,] %>%
  filter(prop<0.6)

plot_6 <- ggplot(data = pop65andabove,
       mapping = aes(fill = pop65andabove$prop)) +
  geom_sf(data = pop65andabove$geometry, color = "gray", lwd = 0.1) + 
  scale_fill_distiller(palette = "YlOrRd", direction = 1) + 
  theme_void() +
  labs(title = "Proportion of population 65+ years of age",
       fill = "") +
  theme(plot.title = element_text(size=8))
```

We will plot all six plots together with the help of gridExtra.
```{r}
library(gridExtra)
grid.arrange(plot_1, plot_2, plot_3, plot_4, plot_5, plot_6, ncol = 3)
```

```{r}
save(data = uninsured, median_income, selfid_whites, households4, commutebybus, pop65andabove, file = "socioeconomic.Rdata")
```
